Higher Moments Distances from Robust Subspace for Botnet Detection

### 2019-05-04 docs/papers/2019JNCA_m-rpca/2019JNCA_m-rpca.tex
working on data model

\begin{abstract}
Distributed attacks organized by botnet has increasing and demanding the development of counter measures in order to detect and avoid botnet's attacks. This paper proposes an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggretation, in a semi-supervisioned fashion and without the computational cost of new robust subspace learning for new observations. Experimental results reveal that the proposed approach outerperforms widely adoted algorithms and a method based on robust estimates and robust skewness for network anomaly detection.
\end{abstract}

@@

Anomaly detection can be defined as the identification of rare and suspicions events by differing the normal or majority of the data. Anomalies are also referred to as outliers, novelties, noise or deviations, and can be related to network attacks, frauds or defects \cite{bhuyan2014network,ahmed2016survey}. Anomalies can be hard to identify and separate from normal data due to the rare occurrencies of anomalies in comparison to normal events, therefore anomaly detection algorithms have to be high discriminative, robust to contamination and able to deal with the imbalanced data problem \cite{he2008learning}.

Several efforts and reasearches aim to avoid network attacks based on known attackers, fingerprints or other patterns. However, distributed attacks organized by botnet has increasing and demanding the development of counter measures in order to detect and avoid unkown attacks or even to deal with adversarial changes on behavior, location and other patterns, in order to avoid detection by behavioral or pattern based detecion systems \cite{gu2008botminer, garcia2014empirical,khattak2015botflex,acarali2016survey,wang2017botnet,Wang2018ddosbotnetssurvey}. To face this problem, it is possible to adopt unsupervised or semi-pervised approaches for network anomaly detection, where it is not necessary known anomalies for trannning models \cite{moustafa2019holistic}.

@@

% Say that the data is lognormal \cite{benson2010network} and link it to skewness https://www.weibull.com/hotwire/issue47/relbasics47.htm
Some widely adopted algorithms assume a gaussian distributed data, however this assumption may not be observed in real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network}. Findings of Benson \emph{et al.}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings show that the skeweness and heavy-tailed distributions can impact algorithms that rely on guassian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network anomaly detection. Therefore, learning methods for imbalanced and skewed data have been atracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}. Skewness and kurtosis are higher moments that can have their measurement as very sensitive for outlier detection, considering that the properties of the kurtosis has been used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

Principal Component Analysis (PCA) is very sensitive to anomalies, thus outliers can corrupt the entire data or can be explored to highlight unespected changes and indicate attacks or frauds \cite{callegari2011novel,Lee2013,vieira2017model}. However, PCA is not precise to reveal flows, time or detailed information for each anomaly, and sometimes relies into visual analysis for pricipal component selection. Robust Principal Component Analysis (RPCA) \cite{candes2011robust} is an extension of PCA that aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, which refers to outliers’ matrix. Therefore, robust subspace learning has receiving a growing atention of researchers aiming the development of network anomlay detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd, hubert2005robpca,hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.

@@

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data. The botnet detection algorithms can also be improved by feature engineering from aggregated traffic, considering that new features can make the data more discriminative for imbalanced learning problems. We also believe that the distance between robust estimates of normal traffic and anomalies can highlight discrepancies and be used for botnet detection. Therefore, we propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggretation. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. The anomaly detection for new observations evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervisioned fashion and without the computational cost of new robust subspace learning for new observations.

We propose a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adoted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. We evaluate the accuracy of the proposed approach for botnet detecion on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statisctical approaches, which are K-means and Gaussian Mixture Model (GMM) \cite{gaddam2007kmeans,moustafa2019holistic}, respectively. We also evaluate the results of ROBPCA \cite{hubert2005robpca}, wich is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness.

@@

% botnet detection
A botnet is a network of bots, that are compromised machines under the influence of malware (bot). The botnet is commandeered by a botmaster and used as resource for attacks, such as distributed denial-of-service (DDoS) attacks, and fraudulent activities such as spam, phishing, identity theft, and information exfiltration. We refer to \cite{ahmed2016survey,moustafa2019holistic} for an overview of network attacks. The botmaster coordenate a botnet through a command and control (C\&C) channel where bots receive commands and syncronize attacks and fraudulent activities. Centralized C\&C structures using the Internet Relay Chat (IRC) protocol have been utilized by botmasters for a long time, but other protocols, such as HTTP, and achitectures, such as Peer-To-Peer, have also been adopted \cite{gu2008botminer}.

@@

% justificative for the dataset and present who else use it
Several approaches for network attack detections uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. However, NSL-KDD dataset maintain the limitations of the KDD 99 regarding volume and reproduction of network traffic similar to the real networks.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by normal and background labeled data, in an imbalanced distribution like in a real network. The autors recommend that none of the botnet families used in the training and crossvalidation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach of \cite{garcia2014empirical}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

% Anomaly detection by robust subspace larning
The problem of PCA or subspace learning for outlier corrupted data is called robust PCA (RPCA) or robust subspace learning \cite{vaswani2018robust}, which can be defined as the decomposition of a given data matrix $\textbf{M}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. This definition is also referred to as the sparse + low rank (S + LR) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning. Robust subspace learning has receiving a growing atention of researchers aiming the development of anomlay detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.
RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adoted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of location and scatter matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness.

@@

% feature engineering and aggregation
Lakhina \emph{et al.} \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Lakhina \emph{et al.} show that the analysis of feature distributions leads to significant advances on sensitive detection of a wide range of anomalies, augmenting detections by volume-based methods, while the anomaly detection approach leads gains on automatic classification of anomalies via unsupervised learning. Goldberg and Shan \cite{goldberg2015importance} argue, based on experience and on putblished real-life examples of anomaly detectors in use at eBay, that processing monitored signals into features is a more fruitful for anomaly detection systems than devising complex statistical tests.

Benson \emph{et al.} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al.} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings highligh that the skeweness and heavy-tailed distributions can impact algorithms that rely on guassian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network traffic analysis.

Considering that learning methods for imbalanced and skewed data have been atracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}, higher moments can be used for outlier detection and discriminative analysis of skewed data. Skewness and the kurtosis are higher moments that can contribute to sensitive outlier detectors, taking into account that the properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection in skewed data sets \cite{hubert2009robustskewed}.

We propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggretation from the CTU-13 data set. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. Thus, we propose that the anomaly detection for new observations should evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervisioned fashion and without the computational cost of new robust subspace learning for new observations.

We propose a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adoted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. Finally, We evaluate the accuracy of the proposed approach for botnet detecion on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

@@

The types of network attacks and C\&C present in CTU-13 data set are:
\begin{itemize}
	\item \textbf{Attacks:} Click Fraud (CF), Port Scan (PS), Compiled and Controled by Authors (US), SPAM and DDOS;
	\item \textbf{C\&C:} IRC, P2P and HTTP.
\end{itemize}

For all the scenarios, the authors of the CTU-13 data set convert the captured pcap files to NetFlows and release the processed flows. The data set contains ground-truth labels for flows as follows: flows from or to the infected machines are labeled as “botnet”; flows from or to noninfected machines are labeled as “normal”; all other flows are labeled as “background.”

The data set orignally contains the following features:  Start Time, End Time, Duration, Source IP Address, Source Port, Direction, Destination IP Address, Destination Port, State, SToS, Total of Packets and Total of Bytes.

Table \ref{tab:tab01} presents an overview organized by scenario and shows the malwares used for botnet attacks, the types of attacks and C\&C types, the total number of flows, the number of malicious flows, which includes the C\&C, and finally shows the number of C\&C and attack flows separatelly.

@@

The CTU-13 aims to reproduce the percentual of C\&C and attack flows according to real cases, where the number of malicious flows are very low in comparison to normal and background flows, resulting into a very imbalanced and challenging data set for botnet detection.

An Exploratory Data Analysis (EDA) reveals that distributions of original features of the CTU-13 are skewed and presents an overlaping between normal and anomalous flows. Following we present the Figure \ref{fig:fig01}, that presents the distribution of flow duration for scenario 12, and exemplifies the data skewness overlaping. We adopt a sampling method based on data aggregation for obtaining more discriminative features, in accordance to important approaches for network anomaly detection \cite{lakhina2005mining, callegari2011novel, garcia2014empirical, chandrashekar2014survey,acarali2016survey, vieira2017model}.

@@

Based on related works \cite{lakhina2005mining,gu2008botminer,callegari2011novel,chandrashekar2014survey,garcia2014empirical,acarali2016survey}, previous experience \cite{vieira2017model, galibus2017offline} and experimental evaluation of our proposal for botnet detection, we selected some features for improving the anomaly detection by means of a method based on higher moments distances from robust subspace. Following we list the proposed features:

\begin{itemize}
	\item \textbf{normal-flow-count:} count of normal (known) flow.
	\item \textbf{background-flow-count :} count of background (unknown) flows.
	\item \textbf{avg-duration:} average duration time of flows.
	\item \textbf{n-dports<1024:} count of destinations ports lower than 1024.
	\item \textbf{n-sports<1024:} count of source ports lower than 1024.
	\item \textbf{n-s-a-p-address:} count of class A source address.	
	\item \textbf{n-s-b-p-address:} count of class B source address.	
	\item \textbf{n-s-c-p-address:} count of class C source address.	
	\item \textbf{n-s-na-p-address:} count of class D* source address.	
\end{itemize}

The features are generated by a data aggregation into a selected window size. We evalute our proposal for windows with 0.15 and 0.25 seconds, for each scenario of the CTU-13 data set.

Figures \ref{fig:fig07} and \ref{fig:fig08} show the distribution of average flow duration, for scenario 12, aggregated by windows of 0.15 and 2 seconds.

In comparison to Figure \ref{fig:fig06}, that shows the distribution of flow duration for the same scenario, we can see that the feature based on the average is more discriminative.

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/agg_distplot_0_15s_12_avg_duration.png}
     \caption{Distribution of average flow duration for scenario 12, aggregated by windows of 0.15 seconds.}
     \label{fig:fig07}
\end{figure}

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/agg_distplot_2s_12_avg_duration.png}
     \caption{Distribution of average flow duration for scenario 12, aggregated by windows of 2 seconds.}
     \label{fig:fig08}
\end{figure}

@@

Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statisctical approaches, which are K-means and Gaussian Mixture Model (GMM) \cite{gaddam2007kmeans,moustafa2019holistic}, respectively. We also evaluate the results of ROBPCA \cite{hubert2005robpca}, wich is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness. In contrast to Garcia \emph{et al.} \cite{garcia2014empirical}, we propose to evaluate each scenario of CTU-13 individually, in a semi-supervised approach that does not rely on training data with anomalies.

@@

	Gargia et al. \cite{garcia2014empirical} propose a division of CTU-13's scearios into train and test, for evaluation of network attack detection algorithms. 
	However, we decide to evaluate our proposed approach for each scenario individually, considering that in the division proposed by Gargia et al. some bots are present only for training and not for testing. 
	Although our proposal does not rely on anomalous data for training, we desire to evaluate our proposal to detect all types of bots captured by CTU-13 data set.

Why GMM?
As clustering-based method, we selected K-means, which is the clustering algorithm most adopted for general purposes and for network anomaly detection \cite{gaddam2007kmeans,moustafa2019holistic}.
It is observed that network traffic do not belong to a Gaussian distribution \cite{benson2010network,moustafa2019holistic}, algorithms based on non-Gaussian distributions, such as GMM, has been adopted for network anomaly detection \cite{moustafa2019holistic}.

@@

Future research is directed to improvements for obtaining better false positive rates, as well as for make the proposed framework able to identify sparse probe attacks or subtle behaviors, such as exfiltration or covert communication, considering the evaluation of a flow-based analysis and novel data sets. Distributed or parallel processing can also be evaluated to analyze the scalability and processing capacity for monitoring high thoughput network traffic. Additionally, future research can evaluate the application of the proposed approach to different attack types and domains, considering cases that are aware to behavioral analysis.

>>>>

\begin{abstract}
Distributed attacks organized by botnet has increasing and demanding the development of counter measures in order to detect and avoid botnet's attacks. This paper proposes an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggregation, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations. Experimental results reveal that the proposed approach outperforms widely adopted algorithms and a method based on robust estimates and robust skewness for network anomaly detection.
\end{abstract}

@@

Anomaly detection can be defined as the identification of rare and suspicions events by differing the normal or majority of the data. Anomalies are also referred to as outliers, novelties, noise or deviations, and can be related to network attacks, frauds or defects \cite{bhuyan2014network,ahmed2016survey}. Anomalies can be hard to identify and separate from normal data due to the rare occurrences of anomalies in comparison to normal events, therefore anomaly detection algorithms have to be high discriminative, robust to contamination and able to deal with the imbalanced data problem \cite{he2008learning}.

Several efforts and researches aim to avoid network attacks based on known attackers, fingerprints or other patterns. However, distributed attacks organized by botnet has increasing and demanding the development of counter measures in order to detect and avoid unknown attacks or even to deal with adversarial changes on behavior, location and other patterns, in order to avoid detection by behavioral or pattern based detection systems \cite{gu2008botminer, garcia2014empirical,khattak2015botflex,acarali2016survey,wang2017botnet,Wang2018ddosbotnetssurvey}. To face this problem, it is possible to adopt unsupervised or semi-supervised approaches for network anomaly detection, where it is not necessary known anomalies for training models \cite{moustafa2019holistic}.

@@

% Say that the data is lognormal \cite{benson2010network} and link it to skewness https://www.weibull.com/hotwire/issue47/relbasics47.htm
Some widely adopted algorithms assume a gaussian distributed data, however this assumption may not be observed in real world problems, such as the case of network traffic analysis, where network traffic features are usually more characterized by skewed and heavy-tailed distributions \cite{lakhina2005mining,benson2010network}. Findings of Benson \emph{et al.}  \cite{benson2010network} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings show that the skeweness and heavy-tailed distributions can impact algorithms that rely on guassian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network anomaly detection. Therefore, learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}. Skewness and kurtosis are higher moments that can have their measurement as very sensitive for outlier detection, considering that the properties of the kurtosis has been used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

Principal Component Analysis (PCA) is very sensitive to anomalies, thus outliers can corrupt the entire data or can be explored to highlight unexpected changes and indicate attacks or frauds \cite{callegari2011novel,Lee2013,vieira2017model}. However, PCA is not precise to reveal flows, time or detailed information for each anomaly, and sometimes relies into visual analysis for principal component selection. Robust Principal Component Analysis (RPCA) \cite{candes2011robust} is an extension of PCA that aims to be resilient to outliers by means of a robust subspace learning \cite{vaswani2018robust} for outlier corrupted data, decomposing a given data matrix $\textbf{X}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, which refers to outliers’ matrix. Therefore, robust subspace learning has receiving a growing attention of researchers aiming the development of network anomaly detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd, hubert2005robpca,hubert2009robustskewed, pascoal2012robust, zhou2017anomaly}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.

@@

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data. The botnet detection algorithms can also be improved by feature engineering from aggregated traffic, considering that new features can make the data more discriminative for imbalanced learning problems. We also believe that the distance between robust estimates of normal traffic and anomalies can highlight discrepancies and be used for botnet detection. Therefore, we propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggregation. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. The anomaly detection for new observations evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations.

We propose a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adopted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. We evaluate the accuracy of the proposed approach for botnet detection on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statistical approaches, which are K-means and Gaussian Mixture Model (GMM) \cite{gaddam2007kmeans,moustafa2019holistic}, respectively. We also evaluate the results of ROBPCA \cite{hubert2005robpca}, which is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness.

@@

% botnet detection
A botnet is a network of bots, that are compromised machines under the influence of malware (bot). The botnet is commandeered by a botmaster and used as resource for attacks, such as distributed denial-of-service (DDoS) attacks, and fraudulent activities such as spam, phishing, identity theft, and information ex-filtration. We refer to \cite{ahmed2016survey,moustafa2019holistic} for an overview of network attacks. The botmaster coordinate a botnet through a command and control (C\&C) channel where bots receive commands and synchronize attacks and fraudulent activities. Centralized C\&C structures using the Internet Relay Chat (IRC) protocol have been utilized by botmasters for a long time, but other protocols, such as HTTP, and architectures, such as Peer-To-Peer, have also been adopted \cite{gu2008botminer}.

@@

% justificative for the dataset and present who else use it
Several approaches for network attack detection uses the KDD 99 \cite{ahmed2016survey,osanaiye2016distributed,bhuyan2014network} data sets for accuracy and performance evaluation, due to their availability and labeled attacks. Even though the KDD 99 data set are criticized by the generation procedure and the risk of over-estimations of anomaly detection due to data redundancy, it still represents one of the few publicly available labeled data sets currently in use today by researchers \cite{osanaiye2016distributed,bhuyan2014network}. NSL-KDD \cite{tavallaee2009detailed} data set is the refined version of the KDD 99 data set that redundant data records are removed, in order to avoid biased classifications. However, NSL-KDD data set maintain the limitations of the KDD 99 regarding volume and reproduction of network traffic similar to the real networks.

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by normal and background labeled data, in an imbalanced distribution like in a real network. The authors recommend that none of the botnet families used in the training and cross-validation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach of \cite{garcia2014empirical}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

% Anomaly detection by robust subspace larning
The problem of PCA or subspace learning for outlier corrupted data is called robust PCA (RPCA) or robust subspace learning \cite{vaswani2018robust}, which can be defined as the decomposition of a given data matrix $\textbf{M}$ into the sum of a low rank matrix $\textbf{L}$, whose column subspace gives the principal components, and a sparse matrix $\textbf{S}$, with outliers or noise. This definition is also referred to as the sparse + low rank (S + LR) formulation \cite{vaswani2018robust}. We refer to \cite{lerman2018overview} and \cite{vaswani2018robust} for more details regarding robust subspace learning. Robust subspace learning has receiving a growing attention of researchers aiming the development of anomaly detection systems \cite{rousseeuw1984mcd, rousseeuw1999fastmcd}, considering outlier-robust methods and sparse-corruption methods \cite{lerman2018overview}.
RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adopted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of location and scatter matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness.

@@

% feature engineering and aggregation
Lakhina \emph{et al.} \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Lakhina \emph{et al.} show that the analysis of feature distributions leads to significant advances on sensitive detection of a wide range of anomalies, augmenting detections by volume-based methods, while the anomaly detection approach leads gains on automatic classification of anomalies via unsupervised learning. Goldberg and Shan \cite{goldberg2015importance} argue, based on experience and on published real-life examples of anomaly detectors in use at eBay, that processing monitored signals into features is a more fruitful for anomaly detection systems than devising complex statistical tests.

Benson \emph{et al.} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al.} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings highlight that the skeweness and heavy-tailed distributions can impact algorithms that rely on gaussian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network traffic analysis.

Considering that learning methods for imbalanced and skewed data have been attracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}, higher moments can be used for outlier detection and discriminative analysis of skewed data. Skewness and the kurtosis are higher moments that can contribute to sensitive outlier detectors, taking into account that the properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection in skewed data sets \cite{hubert2009robustskewed}.

We propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggregation from the CTU-13 data set. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. Thus, we propose that the anomaly detection for new observations should evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervised fashion and without the computational cost of new robust subspace learning for new observations.

We propose a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adopted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. Finally, We evaluate the accuracy of the proposed approach for botnet detecion on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

@@

The types of network attacks and C\&C present in CTU-13 data set are:
\begin{itemize}
	\item \textbf{Attacks:} Click Fraud (CF), Port Scan (PS), Compiled and Controlled by Authors (US), SPAM and DDOS;
	\item \textbf{C\&C:} IRC, P2P and HTTP.
\end{itemize}

For all the scenarios, the authors of the CTU-13 data set convert the captured pcap files to NetFlows and release the processed flows. The data set contains ground-truth labels for flows as follows: flows from or to the infected machines are labeled as “botnet”; flows from or to non infected machines are labeled as “normal”; all other flows are labeled as “background.”

The data set originally contains the following features:  Start Time, End Time, Duration, Source IP Address, Source Port, Direction, Destination IP Address, Destination Port, State, SToS, Total of Packets and Total of Bytes.

Table \ref{tab:tab01} presents an overview organized by scenario and shows the malwares used for botnet attacks, the types of attacks and C\&C types, the total number of flows, the number of malicious flows, which includes the C\&C, and finally shows the number of C\&C and attack flows separately.

@@

The CTU-13 aims to reproduce the percentage of C\&C and attack flows according to real cases, where the number of malicious flows are very low in comparison to normal and background flows, resulting into a very imbalanced and challenging data set for botnet detection.

An Exploratory Data Analysis (EDA) reveals that distributions of original features of the CTU-13 are skewed and presents an overlapping between normal and anomalous flows. Following we present the Figure \ref{fig:fig01}, that presents the distribution of flow duration for scenario 12, and exemplifies the data skewness overlapping. We adopt a sampling method based on data aggregation for obtaining more discriminative features, in accordance to important approaches for network anomaly detection \cite{lakhina2005mining, callegari2011novel, garcia2014empirical, chandrashekar2014survey,acarali2016survey, vieira2017model}.

@@

Based on related works \cite{lakhina2005mining,gu2008botminer,callegari2011novel,chandrashekar2014survey,garcia2014empirical,acarali2016survey}, previous experience \cite{vieira2017model, galibus2017offline} and cross validation  of our proposal for botnet detection, we selected the following features for improving the anomaly detection by means of a method based on higher moments distances from robust subspace.

\begin{itemize}
	\item \textbf{background-flow-count :} Count of unknown flows.
	\item \textbf{normal-flow-count:} Count of known flows.
	\item \textbf{avg-duration:} Average time of flows.
	\item \textbf{n-dports<1024:} Count of destinations ports lower than 1024.
	\item \textbf{n-sports<1024:} Count of source ports lower than 1024.
	\item \textbf{n-s-a-p-address:} Count of class A source address.
	\item \textbf{n-s-b-p-address:} Count of class B source address.
	\item \textbf{n-s-c-p-address:} Count of class C source address.
	\item \textbf{n-s-na-p-address:} Count of source address that are not class A, B or C.
\end{itemize}

Since the selected features are counting and average calculations, the features are generated through a data aggregation by time window. We evaluate time windows of 0.15 and 0.25 seconds, taking into account that lower windows reduce the meantime between the data collection and results of the botnet detection approach.

Figure \ref{fig:fig02} shows the distribution of average flow duration, for scenario 12, aggregated by windows of 0.15 and 2 seconds. In comparison to Figure \ref{fig:fig01}, that shows the distribution of flow duration for the same scenario, we can see that the feature based on the average is more discriminative.

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/agg_distplot_0_15s_12_avg_duration.png}
     \caption{Average flow duration for scenario 12, aggregated by 0.15 seconds.}
     \label{fig:fig02}
\end{figure}

The data is modeled into the matrix $\textbf{M}$ with rows as observations and columns the features.

@@

Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statistical approaches, which are K-means and Gaussian Mixture Model (GMM) \cite{gaddam2007kmeans,moustafa2019holistic}, respectively. We also evaluate the results of ROBPCA \cite{hubert2005robpca}, which is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness. In contrast to Garcia \emph{et al.} \cite{garcia2014empirical}, we propose to evaluate each scenario of CTU-13 individually, in a semi-supervised approach that does not rely on training data with anomalies.

@@

Gargia et al. \cite{garcia2014empirical} propose a division of CTU-13's scenarios into train and test, for evaluation of network attack detection algorithms. 
However, we decide to evaluate our proposed approach for each scenario individually, considering that in the division proposed by Gargia et al. some bots are present only for training and not for testing. 
Although our proposal does not rely on anomalous data for training, we desire to evaluate our proposal to detect all types of bots captured by CTU-13 data set.

Why GMM?

As clustering-based method, we selected K-means, which is the clustering algorithm most adopted for general purposes and for network anomaly detection \cite{gaddam2007kmeans,moustafa2019holistic}.
It is observed that network traffic do not belong to a Gaussian distribution \cite{benson2010network,moustafa2019holistic}, algorithms based on non-Gaussian distributions, such as GMM, has been adopted for network anomaly detection \cite{moustafa2019holistic}.

@@

Future research is directed to improvements for obtaining better false positive rates, as well as for make the proposed framework able to identify sparse probe attacks or subtle behaviors, such as exfiltration or covert communication, considering the evaluation of a flow-based analysis and novel data sets. Distributed or parallel processing can also be evaluated to analyze the scalability and processing capacity for monitoring02       rch can evaluate the application of the proposed approach to different attack types and domains, considering cases that are aware to behavioral analysis.
